---
title: "Getting started with acquire"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with acquire}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Researchers working with external data — from government portals, institutional
repositories, or public APIs — need to document where that data came from. The
`acquire` package downloads files from known URLs and produces a provenance
record alongside the data: SHA-256 hashes, timestamps, HTTP metadata, and a
BibTeX citation.

This vignette walks through the workflow using the USGS earthquake feed.

## Setup

```{r setup}
library(acquire)
```

Set a store path — the directory where `acquire` keeps downloaded files and
their provenance records:

```{r init}
store <- tempfile("quake-store")
acq_store(store)
```

In practice you'd use a permanent path like `"data/raw"` rather than a
temporary directory. The directory is created automatically on first download.

## Define a data source

An `acq_source` object describes where data lives and what it is. The metadata
you provide here flows into the provenance record and BibTeX citation:

```{r source}
usgs_quakes <- acq_source(
  name = "usgs-earthquakes-m4.5",
  landing_url = "https://earthquake.usgs.gov/earthquakes/feed/v1.0/csv.php",
  data_urls = c(
    "earthquakes.csv" = "https://earthquake.usgs.gov/earthquakes/feed/v1.0/summary/4.5_month.csv"
  ),
  title = "USGS M4.5+ Earthquakes, Past 30 Days",
  publisher = "U.S. Geological Survey",
  year = format(Sys.Date(), "%Y"),
  description = "Magnitude 4.5+ earthquakes from the past 30 days, updated every minute."
)

usgs_quakes
```

Named entries in `data_urls` control the local filename. Unnamed entries use
the URL basename.

## Download and document

`acq_download()` fetches all data files, computes SHA-256 hashes, records HTTP
metadata (ETag, Last-Modified, Content-Type), and writes a provenance record:

```{r download}
acq_download(usgs_quakes)
```

The source directory now contains the data alongside its provenance record:

```{r files}
list.files(file.path(store, "usgs-earthquakes-m4-5"), recursive = TRUE, all.files = TRUE, no.. = TRUE)
```

## The provenance record

The heart of the package is `provenance.json` — a structured record of
exactly what was downloaded, from where, and when. You can read it as a list:

```{r provenance}
prov <- acq_read_provenance(usgs_quakes)
prov$files[["earthquakes.csv"]][c("url", "downloaded", "size", "sha256")]
```

This record is written at download time and never modified unless you
re-download the source. The SHA-256 hash proves the file hasn't changed since
acquisition.

## Verify integrity

`acq_verify()` re-computes hashes and compares them to the recorded values.
This catches accidental edits, corruption, or file moves:

```{r verify}
acq_verify(usgs_quakes)
```

## Citation

`acq_download()` automatically generates a BibTeX citation in
`data-sources.bib`. You can also regenerate it manually:

```{r cite}
acq_cite(usgs_quakes)
```

The resulting `data-sources.bib` can be used directly with R Markdown, Quarto,
or LaTeX.

## Status overview

`acq_status()` gives a summary of all sources in the store:

```{r status}
acq_status()
```

```{r cleanup, include = FALSE}
unlink(store, recursive = TRUE)
```
