---
title: "Getting started with acquire"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with acquire}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Researchers working with external data — from government portals, institutional
repositories, or public APIs — need to document where that data came from. The
`acquire` package downloads files from known URLs or registers local files, and
produces a provenance record alongside the data: SHA-256 hashes, timestamps,
HTTP metadata, and a BibTeX citation.

This vignette walks through both workflows: downloading remote data from the
USGS earthquake feed, and registering local files that arrive outside of a URL.

## Setup

```{r setup}
library(acquire)
```

Set a store path — the directory where `acquire` keeps downloaded files and
their provenance records:

```{r init}
store <- tempfile("quake-store")
acq_store(store)
```

In practice you'd use a permanent path like `"data/raw"` rather than a
temporary directory. The directory is created automatically on first download.

## Define a data source

An `acq_source` object describes where data lives and what it is. The metadata
you provide here flows into the provenance record and BibTeX citation:

```{r source}
usgs_quakes <- acq_source(
  name = "usgs-earthquakes-m4.5",
  landing_url = "https://earthquake.usgs.gov/earthquakes/feed/v1.0/csv.php",
  data_urls = c(
    "earthquakes.csv" = "https://earthquake.usgs.gov/earthquakes/feed/v1.0/summary/4.5_month.csv"
  ),
  title = "USGS M4.5+ Earthquakes, Past 30 Days",
  publisher = "U.S. Geological Survey",
  year = format(Sys.Date(), "%Y"),
  description = "Magnitude 4.5+ earthquakes from the past 30 days, updated every minute."
)

usgs_quakes
```

Named entries in `data_urls` control the local filename. Unnamed entries use
the URL basename.

## Download and document

`acq_download()` fetches all data files, computes SHA-256 hashes, records HTTP
metadata (ETag, Last-Modified, Content-Type), and writes a provenance record:

```{r download}
acq_download(usgs_quakes)
```

The source directory now contains the data alongside its provenance record:

```{r files}
list.files(file.path(store, "usgs-earthquakes-m4-5"), recursive = TRUE, all.files = TRUE, no.. = TRUE)
```

## The provenance record

The heart of the package is `provenance.json` — a structured record of
exactly what was downloaded, from where, and when. You can read it as a list:

```{r provenance}
prov <- acq_read_provenance(usgs_quakes)
prov$files[["earthquakes.csv"]][c("url", "downloaded", "size", "sha256")]
```

This record is written at download time and updated when you refresh the
source. The SHA-256 hash proves the file hasn't changed since acquisition.

## Verify integrity

`acq_verify()` re-computes hashes and compares them to the recorded values.
This catches accidental edits, corruption, or file moves:

```{r verify}
acq_verify(usgs_quakes)
```

## Citation

`acq_download()` automatically generates a BibTeX citation in
`data-sources.bib`. You can also regenerate it manually:

```{r cite}
acq_cite(usgs_quakes)
```

The resulting `data-sources.bib` can be used directly with R Markdown, Quarto,
or LaTeX.

## Refreshing a source

External data changes over time. `acq_refresh()` re-downloads each file,
compares hashes to the provenance record, and updates the local store only if
something has changed. Before updating, it archives the current files in a
timestamped subdirectory:

```{r refresh}
acq_refresh(usgs_quakes)
```

If nothing has changed, the function reports that and returns early — no files
are modified. Note that `acq_download()` is strictly for first-time
acquisition; calling it again on an existing source will error and recommend
`acq_refresh()`.

## Status overview

`acq_status()` gives a summary of all sources in the store:

```{r status}
acq_status()
```

## Registering local files

Not all data arrives via URL. Files received by email, copied from a shared
drive, or exported from a database can be registered with `acq_register()`.
This gives them the same provenance tracking as downloaded files.

First, let's create some example files to register:

```{r local-setup}
# Simulate local files (in practice these already exist somewhere)
local_dir <- tempfile("local-data")
dir.create(local_dir)
writeLines("id,value\n1,10\n2,20\n3,30", file.path(local_dir, "survey.csv"))
writeLines("id: participant ID\nvalue: response score", file.path(local_dir, "codebook.txt"))
```

Define a source using `data_paths` and `metadata_paths` instead of URLs:

```{r local-source}
survey <- acq_source(
  name = "annual-survey",
  data_paths = c("survey.csv" = file.path(local_dir, "survey.csv")),
  metadata_paths = c("codebook.txt" = file.path(local_dir, "codebook.txt")),
  title = "Annual Participant Survey",
  publisher = "Research Lab",
  year = format(Sys.Date(), "%Y")
)

survey
```

A source must use _either_ URLs or paths — mixing them is an error.

### Register into the store

`acq_register()` copies files into the canonical store structure, computes
hashes, and writes a provenance record:

```{r register}
acq_register(survey)
```

The store structure mirrors what `acq_download()` creates:

```{r local-files}
list.files(file.path(store, "annual-survey"), recursive = TRUE, all.files = TRUE, no.. = TRUE)
```

### Move instead of copy

If you don't need the originals, use `move = TRUE` to avoid duplication. The
original is deleted only after the copy is verified:
    
```r
acq_register(source, move = TRUE)
```

### Registering files already in place

If you've already placed files in the correct store location, `acq_register()`
detects this and skips the copy — it just hashes and records provenance.

### Checking for upstream changes

The same functions work for both remote and local sources:

```{r local-verify}
acq_verify(survey)
```

`acq_check()` compares file size and modification time against the original
source path (the equivalent of an HTTP HEAD check for remote files):

```{r local-check}
acq_check(survey)
```

If the original file has changed, you can use `acq_compare()` for a definitive
hash-based comparison, and `acq_refresh()` to update the store:

```{r local-compare}
acq_compare(survey)
```

### Status overview (revisited)

`acq_status()` shows both remote and local sources, with an `origin` column:

```{r status-both}
acq_status()
```

```{r cleanup, include = FALSE}
unlink(store, recursive = TRUE)
unlink(local_dir, recursive = TRUE)
```
